---
title: "hcocena: a complete worked analysis"
package: hcocena
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{hcocena: a complete worked analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
# BiocStyle trims figure margins with ImageMagick; several plots here are drawn
# by functions that manage their own graphics devices, which can leave the hook
# a figure it cannot compute a bounding box for.
knitr::opts_chunk$set(crop = NULL)
```

# What hcocena does

`hcocena` builds **co-expression networks** and reads biology off their
structure. The idea is that genes which vary together across samples tend to
belong to the same programme, so a gene-gene correlation network, cut into
communities, gives you a handful of *modules* instead of thousands of
individual genes. Each module can then be characterised -- which conditions
drive it, which functions it is enriched for, which genes sit at its centre.

What makes it *h*CoCena ("horizontal CoCena") is that several datasets can be
analysed **as one network**. Different tissues, platforms or cohorts each
contribute their own correlation network; those are merged, and the modules
are found on the merged graph. A module is therefore a programme that exists
across layers, and its behaviour can be compared between them.

This vignette runs the whole analysis end to end on a small simulated dataset,
so every number and figure below is produced when the vignette is built. The
steps that need an external service -- ChEA3, Enrichr, DoRothEA, Cytoscape, an
LLM endpoint -- cannot run here; see [Beyond this vignette](#beyond) for the
full templates that cover them.

# Installation

`hcocena` is distributed through Bioconductor and is installed with
`BiocManager`:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("hcocena")
```

Several optional steps rely on packages listed under `Suggests` (for example
`limma`, `decoupleR`, `RCy3` or `ellmer`); the functions that need them say so
and name the package to install.

```{r libraries, message = FALSE}
library(hcocena)
library(MultiAssayExperiment)
library(SummarizedExperiment)
```

# The example data

Two layers of 40 genes and 32 samples each. The genes are real marker symbols
for four blood cell types, and the simulation makes each set co-vary as a
block, so the analysis should recover four modules that correspond to the four
cell types. That is the point of using a simulation here: we know the answer
in advance and can check that the pipeline finds it.

The samples are 16 donors measured at two timepoints, split across three
conditions, with a batch label and an age covariate -- enough structure for
the statistics later on. `inst/scripts/make-fixtures.R` in the installed
package contains the generating code.

```{r peek-data}
extdir <- normalizePath(system.file("extdata", package = "hcocena"), winslash = "/")
counts_preview <- read.delim(file.path(extdir, "toy_layer1_counts.tsv"), nrows = 3)
counts_preview[, 1:5]

anno_preview <- read.delim(file.path(extdir, "toy_layer1_anno.tsv"))
head(anno_preview, 3)
```

# Setting up the analysis

Every analysis starts from an empty `HCoCenaExperiment` and is configured by
passing it through a series of functions. The object carries the data, the
settings and every intermediate result, so there is no hidden state to keep
track of.

```{r init}
outdir <- file.path(tempdir(), "hcocena-vignette-output")
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

hc <- hc_init()
hc <- hc_set_paths(
  hc,
  dir_count_data = paste0(extdir, "/"),
  dir_annotation = paste0(extdir, "/"),
  dir_reference_files = paste0(extdir, "/"),
  dir_output = outdir
)
hc <- hc_init_save_folder(hc, name = "")
```

`hc_define_layers()` names the layers and says which files belong to them. The
names are used in file names and plot labels from here on.

```{r layers}
hc <- hc_define_layers(
  hc,
  data_sets = list(
    Layer1 = c("toy_layer1_counts.tsv", "toy_layer1_anno.tsv"),
    Layer2 = c("toy_layer2_counts.tsv", "toy_layer2_anno.tsv")
  )
)

hc <- hc_read_data(
  hc,
  sep_counts = "\t",
  sep_anno = "\t",
  gene_symbol_col = "SYMBOL",
  sample_col = "SampleID",
  count_has_rn = FALSE,
  anno_has_rn = FALSE,
  auto_setup_output = FALSE
)
```

## Global settings

`variable_of_interest` is the annotation column that defines the groups
everything is later compared across; `control_keyword` names the reference
level. `range_GFC` caps the group-fold-change scale introduced below.

```{r global-settings}
hc <- hc_set_global_settings(
  hc,
  organism = "human",
  control_keyword = "control",
  variable_of_interest = "group",
  min_nodes_number_for_network = 1,
  min_nodes_number_for_cluster = 1,
  range_GFC = 2,
  layout_algorithm = "layout_with_stress",
  data_in_log = FALSE
)
```

## Layer settings

`top_var` is the number of most variable genes per layer that enter the
correlation. Genes that barely move across samples carry no co-expression
signal but do add noise and computation, so the network is built on the
variable ones. In a real analysis this is typically a few thousand; here the
dataset only has 40 genes, so we take all of them.

`min_corr` and `range_cutoff_length` define the grid of correlation cutoffs
that the next step scans.

```{r layer-settings}
hc <- hc_set_layer_settings(
  hc,
  top_var = c(40, 40),
  min_corr = c(0.5, 0.5),
  range_cutoff_length = c(50, 50),
  print_distribution_plots = c(FALSE, FALSE)
)
```

# Part I: correlation and choosing a cutoff

`hc_run_expression_analysis_1()` computes all pairwise gene-gene correlations
per layer and then, for every candidate cutoff on the grid, records what the
resulting network would look like.

```{r part1, results = "hide", message = FALSE}
hc <- hc_run_expression_analysis_1(hc, corr_method = "pearson")
```

**Why a cutoff has to be chosen at all.** A correlation matrix is a complete
graph -- every gene is connected to every other one, just with different
weights. To get a network with structure you keep only edges above some
correlation. Set it too low and everything is connected to everything, so
there are no modules to find; set it too high and the network falls apart into
fragments and you lose most of your genes.

`hcocena` picks the cutoff by two criteria at once:

- **Scale-free topology (R²).** Biological networks tend to have few
  highly connected hubs and many sparsely connected genes, a degree
  distribution that follows a power law. R² measures how well the network at a
  given cutoff fits that expectation. Higher is more biologically plausible.
- **How much network is left.** The number of remaining nodes and the mean
  degree. A cutoff with a beautiful R² over eleven surviving genes is useless.

```{r cutoff, results = "hide", message = FALSE}
hc <- hc_set_cutoff(hc, auto = TRUE)
```

The statistics behind the choice can be inspected per layer -- R², node count
and mean degree along the cutoff grid:

```{r cutoff-plot, fig.height = 3.5}
hc_plot_cutoffs(hc, interactive = FALSE)
```

# Part II: group fold changes

`hc_run_expression_analysis_2()` filters each layer's network to the chosen
cutoff and computes **group fold changes** (GFCs).

A GFC summarises, for one gene and one group of samples, how far that group
sits from the overall mean of the gene, on a scale capped at `range_GFC`. It
is what makes modules comparable across layers: absolute expression differs
between platforms and tissues, but "up in severe relative to this gene's own
average" is the same quantity everywhere.

```{r part2, results = "hide", message = FALSE}
hc <- hc_run_expression_analysis_2(hc, plot_HM = FALSE)
```

# Integration

Each layer now has its own filtered network. `hc_build_integrated_network()`
merges them into one graph.

- `mode = "u"` (union) keeps an edge present in *any* layer -- the inclusive
  choice, and the usual one.
- `mode = "i"` (intersection) keeps only edges present in *every* layer, which
  is stricter and needs the layers to measure largely the same genes.
- `multi_edges` says what to do when an edge exists in several layers with
  different weights: `"min"` is conservative, `"mean"` and `"max"` are the
  alternatives.

```{r integrate, results = "hide", message = FALSE}
hc <- hc_build_integrated_network(hc, mode = "u", multi_edges = "min")
```

```{r network-size}
g <- hc_graph(hc)
c(nodes = igraph::vcount(g), edges = igraph::ecount(g))
```

# Module detection

Communities are found with the Leiden algorithm. `resolution` controls
granularity: higher values give more, smaller modules.

Leiden is stochastic, so `hcocena` runs it `no_of_iterations` times and keeps
the consensus. A gene ends up in the module it was assigned to most often, and
genes that never settle are dropped -- which is why a module assignment from
this package is more stable than a single Leiden run.

```{r cluster, results = "hide", message = FALSE}
hc <- hc_cluster_calculation(
  hc,
  cluster_algo = "cluster_leiden",
  no_of_iterations = 5,
  resolution = 1
)
```

```{r cluster-sizes}
gtc <- hc_gene_to_cluster(hc)
table(gtc$color)
```

Four modules of ten genes, as the simulation intended. At this point the
modules are identified by colour; the readable `M1`-style labels are assigned
when the module heatmap is drawn, which also shows their GFC profile across
the groups of both layers:

```{r heatmap, fig.height = 4.5, results = "hide", message = FALSE}
hc <- hc_plot_cluster_heatmap(hc, file_name = FALSE)
```

Module membership per gene is the table most downstream steps work from:

```{r module-genes}
module_genes <- as.data.frame(hc_satellite(hc, "module_gene_list"))
split(module_genes$genes, module_genes$module)[["M1"]]
```

# Functional enrichment

Modules are tested for over-representation of gene sets. In a normal analysis
these come from GO, KEGG, Hallmark or Reactome via `gene_sets =`; here we
write a small GMT of the four marker panels instead, so that the vignette
stays offline and self-contained.

**The background matters.** `universe = "all_genes"` (the default) tests each
module against every measured gene, answering "is this module enriched
relative to the transcriptome?". `universe = "network"` tests against the
genes that entered the network -- the genes that could have landed in a module
at all -- answering "is this module enriched relative to the other modules?".
The second removes the bias that top-variance selection and the correlation
cutoff introduce, and is the stricter question.

```{r enrichment, results = "hide", message = FALSE}
gmt <- file.path(tempdir(), "celltype_markers.gmt")
writeLines(c(
  paste(c("T cells", "-", "CD3D", "CD3E", "CD3G", "CD2", "CD28",
          "LCK", "ZAP70", "IL7R", "CD7", "TRAC"), collapse = "\t"),
  paste(c("B cells", "-", "CD19", "MS4A1", "CD79A", "CD79B", "BLNK",
          "PAX5", "CR2", "FCRL1", "TNFRSF13B", "VPREB3"), collapse = "\t"),
  paste(c("Monocytes", "-", "CD14", "LYZ", "FCN1", "VCAN", "S100A8",
          "S100A9", "CSF1R", "ITGAM", "CD68", "FCGR3A"), collapse = "\t"),
  paste(c("NK cells", "-", "NKG7", "GNLY", "KLRD1", "KLRF1", "PRF1",
          "GZMB", "NCR1", "KLRC1", "FGFBP2", "SPON2"), collapse = "\t")
), gmt)

hc <- hc_functional_enrichment(
  hc,
  gene_sets = character(0),
  custom_gmt_files = c(CellTypes = gmt),
  universe = "network"
)
```

```{r enrichment-result}
enr <- hc_satellite(hc, "enrichments")$significant_enrichments_all_dbs
as.data.frame(enr)[, c("module_label", "term", "GeneRatio", "qvalue")]
```

Each module recovers exactly one cell type with all ten of its markers, which
is the answer the simulation was built to have.

# Characterising the modules

## Which conditions drive a module

`hc_module_condition_significance()` reduces each module to one value per
sample -- the mean expression of its genes -- and tests that against the
conditions. Wilcoxon (or Kruskal-Wallis for more than two groups) is
rank-based, `limma` fits a linear model with pairwise contrasts, and an
optional mixed model handles repeated measures via a donor random intercept.

`standardize_modules = TRUE` z-scores each module within a layer first. Raw
means carry the module's absolute expression level, which distorts limma's
variance moderation and makes effect sizes incomparable between modules.
Applied per layer it also removes the offset between layers, which matters
because the model formula itself carries no layer term.

```{r module-significance, results = "hide", message = FALSE}
hc <- hc_module_condition_significance(
  hc,
  condition_col = "group",
  run_limma = FALSE,
  standardize_modules = TRUE,
  export_excel = FALSE
)
```

```{r module-significance-result}
sig <- hc_satellite(hc, "module_condition_significance")$wilcox
as.data.frame(sig)[, c("cluster", "test", "p", "p_adj")]
```

Nothing is significant here, and that is the correct answer: the simulation
assigned the condition labels at random, so there is no group effect to find.

## Correlating modules with metadata

Numeric covariates are correlated with module expression directly; categorical
ones are correlated across the groups of the variable of interest, which needs
at least three of them.

```{r meta-correlation, fig.height = 3.5, results = "hide", message = FALSE}
hc_meta_correlation_num(hc, set = 1, meta = "age")
```

## Module GFC tables

The numbers behind the module heatmap, per group and per sample:

```{r gfc-tables, results = "hide", message = FALSE}
hc <- hc_module_gfc_tables(hc, export_xlsx = FALSE)
```

```{r gfc-tables-result}
head(as.data.frame(hc_satellite(hc, "module_gfc_tables")$by_group), 4)
```

## Hub genes

Hubs are the genes most central to their own module, by weighted degree,
closeness and betweenness. They are the natural candidates when a module has
to be reduced to a few representatives.

```{r hubs, results = "hide", message = FALSE, fig.show = "hide"}
hc <- hc_find_hubs(hc, top = 5)
```

```{r hubs-result}
vapply(hc_satellite(hc, "hub_out"), function(x) paste(x[[1]], collapse = ", "),
       character(1))
```

The hubs of each module are exactly the markers of the corresponding cell
type, which is another way of seeing that the modules are clean.

## Module scores

A module score is the fraction of a gene's neighbours that lie inside its own
module -- a measure of how cleanly the community was separated. Scores near 1
mean the module is well isolated in the graph.

```{r module-scores, fig.height = 3.5, results = "hide", message = FALSE}
hc <- hc_get_module_scores(hc)
```

# The object that carries all of this

Everything above is stored in the same `HCoCenaExperiment`. The assays live in
a `MultiAssayExperiment`:

```{r container}
names(experiments(hc_mae(hc)))
vapply(experiments(hc_mae(hc)), function(x) dim(x)[[1]], integer(1))
```

Settings stay attached, so an analysis can be reconstructed from the object
alone:

```{r config}
as.data.frame(methods::slot(hc_config(hc), "global"))[, c("organism", "voi", "range_GFC")]
```

Downstream results are collected in the satellite slot, which `hc_satellite()`
returns:

```{r satellite}
names(hc_satellite(hc))
```

# Beyond this vignette {#beyond}

Several parts of `hcocena` need services that cannot be reached while a
vignette is being built, and are therefore not shown above:

- transcription-factor enrichment through **ChEA3**
  (`hc_tf_overrep_module()`, `hc_check_tf()`),
- cell-type annotation from **Enrichr** libraries
  (`hc_celltype_annotation()`, `hc_celltype_activity_decoupler()`),
- upstream regulator and pathway inference via **DoRothEA/PROGENy**
  (`hc_upstream_inference()`),
- network layout round-trips with **Cytoscape**
  (`hc_export_to_cytoscape()`),
- LLM-based module interpretation (`hc_module_function_llm()`), optionally
  grounded in retrieved literature and in the enrichment terms computed above.

The complete walkthroughs are installed with the package as R Markdown
templates:

```{r templates}
dir(system.file("scripts", "workflows", package = "hcocena"), pattern = "[.]Rmd$")
```

`hcocena_main.Rmd` is the full analysis; `hcocena_satellite.Rmd` collects the
optional downstream steps. Copy one next to your data and edit the paths at
the top:

```{r copy-template, eval = FALSE}
file.copy(
  system.file("scripts", "workflows", "hcocena_main.Rmd", package = "hcocena"),
  "hcocena_main.Rmd"
)
```

They are templates rather than reproducible documents: they point at your own
count and annotation files and contact the services listed above.

# Session information

```{r session-info}
sessionInfo()
```
